Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services

### 2024-10-10 manuscript/02-literature.qmd
Revised literature review

Precise sales prediction is an essential and inexpensive way for each company to augment their profits, decrease their costs, and achieve greater flexibility to changes. In other words, exact sales forecasting is utilized for capturing the tradeoff between customer demand satisfaction and inventory costs [@gupta2000mid]. Especially, for the pharmaceutical industry, successful sales forecasting systems can be very beneficial, due to the short shelf-life of many pharmaceutical products
and the importance of the product quality which is closely related to the human health [@doganis2006time].There are many decisions that depend on the quality of forecasts in the health care system, from capacity planning to layout decisions to the daily schedules. In general,the role of forecasting in health care is to inform both clinical and non-clinical decisions. While the former concerns decisions related to patients and their treatments [@makridakis2019forecasting], the latter involves policy/management, and supply chain decisions that support the delivery of high-quality care for patients.In the coming paragarphs we will review what other researchers had done in the area of pharmaceuticals forecasting.Morover today,manufacturing companies are organized following the principles of just-in-time (JIT) delivery which focuses on efficiency and eliminating the waste of excessive inventory [@van2021using]. In JIT systems, accurate Stock Keeping Unit (SKU) short-term demand forecasts, typically on a daily, weekly or monthly basis, are crucial as they are the fundamental inputs to drive the supply chain planning system.Therefore, forecasting performance can strongly affect a company’s financial results, since both under- and overestimating short-term demand are expensive as they result in supply shortages and consequently poor customer service, and excess inventories and product obsolescence, respectively (Sanders and Graman, 2009).In this section, we provide a brief review of studies on forecasting harmaceuticals products.In conclusion, the research findings indicated that the hybrid neural network approach, which combines linear and nonlinear modeling techniques, outperformed traditional methods like ARIMA and standalone neural networks in forecasting pharmaceutical sales. The hybrid approach demonstrated the ability to capture both linear and nonlinear patterns in sales data accurately, especially in scenarios with limited historical records for individual drugs.

A study by Khalil Zadeh et al. developed an intelligent sales prediction model for Pharmaceutical Distribution Companies (PDCs) by combining network analysis tools and time series forecasting methods. This approach addresses the challenge of limited historical sales data for pharmaceutical products, making traditional forecasting methods inadequate[@zhang2004neural].The researchers employed Mean Squared Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of various forecasting models, including ARIMA, neural networks, and a hybrid approach combining both. The hybrid model, inspired by Zhang's work[@zhang2004neural], used ARIMA for linear components and neural networks for nonlinear components.Three methods were used: (1) ARIMA, (2) a hybrid neural network with each drug’s past records, and (3) a hybrid neural network with both each drug’s and its group members’ past records. Grouping products and using group members' records significantly improved forecasting accuracy by addressing the issue of limited data and leveraging both linear and nonlinear modeling capabilities.The study concluded that the hybrid approach, incorporating both individual and group past records, performed better than using either method alone [@khalil2014intelligent].

A study by Van Bell et al. used various forecasting models for point of sale data to assess the impact of including sell-through data on forecast accuracy. They compared information sharing (IS) methods with non-information sharing (NIS) methods using the following approaches. Substitution Approach (ETS-W and ARIMA-W): These models use downstream wholesaler sales time series as input instead of prior manufacturer shipments, suitable when all goods pass through wholesalers. Integration Approach (Extrapolative Time Series Methods with Exogenous Variables): This includes methods that integrate downstream information using extrapolative time series techniques and machine learning (ML) approaches. Extrapolative Time Series Methods (ETS and ARIMA): Traditional forecasting methods based on modeling past time series structures and extrapolating into the future. The study evaluated these methods using Mean Squared Error (MSE) and Mean Absolute Error (MAE). Additionally, the researchers used naive methods, IS and NIS with ETS and ARIMA models, and machine learning techniques like LASSO, Support Vector Regression (SVR), and Random Forest (RF).LASSO (Least Absolute Shrinkage and Selection Operator) was identified as the best-performing method for improving forecast accuracy, particularly for one-step ahead forecasts. It outperformed other methods for longer horizons (h > 1) by effectively selecting the most relevant variables and reducing overfitting. ETS was found to be the most effective univariate baseline method, and SVR showed the best performance for h = 1. The study concluded that IS methods generally outperformed NIS methods, indicating that incorporating downstream information, such as sell-through data, can significantly enhance short-term forecast accuracy in supply chain management. Machine learning techniques like LASSO, SVR, and RF faced challenges related to overfitting and model complexity, especially with high-dimensional data. Balancing model complexity and interpretability was crucial for robust forecasting. Overall, the research suggests that LASSO, along with ETS and SVR, are effective for improving demand prediction accuracy in multi-echelon supply chains by leveraging downstream information and collaborative approaches[@van2021using].

A study by Nikolopoulos et al. forecasted UK pharmaceutical time series before and after patent expiry, a critical point when a generic version enters the market. The study used various models, including naïve and moving averages, the Bass diffusion model, repeat purchase diffusion models, three regression models, and two exponential smoothing (ES) approaches for non-trended (simple ES) and trended data (Holt ES) [@nikolopoulos2016forecasting].Holt Exponential Smoothing: This model was the most accurate for forecasting both branded and generic drugs across short-term, mid-term, and long-term horizons.Naïve Model: Used as a benchmark, the relative absolute error (RAE) was calculated to compare forecasting accuracy.Bass Diffusion Model: Although popular in marketing research, it had limitations in accurately forecasting pharmaceutical time series due to the complex dynamics between branded and generic drugs. Comparative Analysis: The study compared various forecasting models to evaluate their performance, highlighting the importance of selecting appropriate models based on the data and context.The research found that Holt’s exponential smoothing model outperformed others, including naïve models, moving averages, and ARIMA models, especially for short-term forecasts. However, forecasting accuracy varied across different horizons. Despite their limitations, diffusion models were included to provide additional insights.Overall, the study emphasized the strengths and limitations of different forecasting approaches in the context of pharmaceutical time series, with Holt’s exponential smoothing model demonstrating superior performance[@nikolopoulos2016forecasting].

Astudy by Siddiqui et al used to develop a combined forecasting model.The combination is developed using two frequent, but separately used models, i.e. Autoregressive Integrated Moving Average (ARIMA)  model and Holts Winter model and a combined model were proposed which is known as the ARHOW model.The ARIMA model is commonly used for time series forecasting, while Holt’s Winter model is effective for handling seasonality in data [@inproceedings{markovska2016comparative].The ARHOW model aims to enhance forecasting accuracy by leveraging the strengths of both ARIMA and Holt’s Winter models. The model equation for obtaining a consolidated forecast is as follows: ARHOW = β * ArimaForecast + β1 * HoltWinterForecast
In this equation:
ARHOW represents the Combined Weighted Forecast. ArimaForecast and HoltWinterForecast are the forecasts obtained by applying the ARIMA and Holt’s Winter models over the training set.β and β1 are the weights assigned to the ARIMA and Holt’s Winter forecasts, respectively. By combining the forecasts generated by these two models, the ARHOW model seeks to mitigate the impact of volatility and external factors on forecasting accuracy, ultimately providing more reliable predictions for pharmaceutical sales data [@siddiqui2022hybrid] .  Combining multiple forecasting techniques like ARIMA and Holt’s Winter to create the ARHOW model introduces complexity. Managing and interpreting the results of a hybrid model may pose challenges in practical implementation and decision-making processes .Morover External factors such as regulatory changes, technological advancements, and market dynamics can introduce forecasting errors. While the ARHOW model aims to improve accuracy, it may still be susceptible to errors resulting from unforeseen external influences [@siddiqui2022hybrid].The key forecasting metrics employed in the study include:

Mean Absolute Percentage Error (MAPE), Mean, Mean Absolute Deviation (MAD), Mean Square Error (MSE), Root Mean Square Error (RMSE).The Mean Absolute Percentage Error (MAPE) and other forecasting metrics suggest that the ARHOW model provides more accurate predictions compared to traditional forecasting techniques like ARIMA, Holt’s Winter, ETS, Theta, and Naïve models. The model's efficiency in forecasting pharmaceutical demand can help manufacturers plan inventory and prevent stockouts or overstock situations .In conclusion, the study underscores the effectiveness of the ARHOW model in improving demand forecasting accuracy in the pharmaceutical industry and provides insights for enhancing supply chain efficiency through advanced forecasting techniques .

A study by de Oliveira et al. compared five machine learning regression approaches for lead time forecasting in the pharmaceutical supply chain. The models used were k-nearest neighbors (KNN), support vector machines (SVM), random forests (RF), linear regression (LR), and multi-layer perceptrons (MLP). These models aimed to predict supplier delivery times, enhancing planning for production and logistics services in the pharmaceutical field [@de2021lead]. The study focused on using machine learning techniques to forecast lead times for purchasing products in the pharmaceutical industry. The comparison aimed to identify the most effective approach for predicting purchasing lead times, demonstrating high accuracy potential, especially with linear support vector regression. Key limitations noted included computational complexity, sensitivity to outliers, difficulty in interpretability, and overfitting. Using mean squared error (MSE) as the forecast accuracy metric, the study found that the SVM approach achieved the best performance, with an average error of less than two days. This indicates that SVM was the most effective model among those tested for forecasting purchasing lead times in the pharmaceutical supply chain[@inproceedings{de2021lead].

A study by Rathipriya et al. focused on demand forecasting for time-series pharmaceutical data using both shallow and deep neural network models[@rathipriya2023demand]. The researchers employed several models, including the Autoregressive Integrated Moving Average (ARIMA) model, known for its simplicity and ability to handle non-stationary series.The shallow neural network models used were, Radial Basis Function Neural Network (RBF_NN),Generalized Regression Neural Network (GR_NN),Probabilistic Neural Network (P_NN),For deep neural networks, the study investigated:Long Short-Term Memory (LSTM),Stacked LSTM,Additionally, the study explored hybrid models, combining ARIMA with the Adaptive Neural Fuzzy Inference System (ANFIS) to forecast Iran's annual energy usage, demonstrating higher accuracy compared to individual models. The researchers also employed methods like Genetic Algorithms (GA), Bayesian structures, and Reinforcing Artificial Neural Networks (RANN) for forecasting short-run demand.The performance of the demand forecast models (DFMs) was evaluated using Root Mean Squared Error (RMSE). The study found that shallow neural network models, such as GR_NN, RBF_NN, and P_NN, had almost identical mean RMSE values across all eight drug categories, indicating their effectiveness in predicting pharmaceutical demand. Conversely, deep neural network models like LSTM and Stacked LSTM did not perform as well due to the limited number of samples available for training. The research concluded that shallow neural network-based DFMs performed well for five out of eight drug categories, while the ARIMA model was best for the remaining three. This suggests that no single forecasting model is optimal for all drug categories[@article{rathipriya2023demand].

A study by Candan et al. on demand forecasting in the pharmaceutical industry utilized a neuro-fuzzy approach as the forecasting method[@inproceedings{hrupelj2017demand]. Neuro-fuzzy modeling combines information from empirical models, heuristics, and data, providing effective models that explain solutions better than black-box models like neural networks[@abraham2001neuro; @babuvska2003neuro]. The study also explored the emotional learning fuzzy inference system (ELFIS) and locally linear neuro-fuzzy (LLNF) methods proposed by Abdollahzade et al. (2012) for predicting the stock prices of a pharmaceutical company[@abdollahzade2012application; @hrupelj2017demand]. While the neuro-fuzzy approach combines fuzzy logic and neural networks to improve forecasting, it may still struggle with extremely complex or non-linear relationships within the data. The study used paired t-tests and mean difference analysis to evaluate forecast accuracy. Overall, the metrics supported the conclusion that the neuro-fuzzy approach is a robust method for demand forecasting in the pharmaceutical industry, providing accurate predictions closely aligned with actual demand quantities.

A study by Burinskiene discusses various forecasting techniques in the pharmaceutical retail industry. These include: Single Exponential Smoothing: Applies weighted averages with exponentially decreasing weights for older observations. Moving Average: Used for trend-based estimation to predict demand and standard deviation.
Naive Method: Assumes all predictions are equal to the last observation, serving as a benchmark. Holt's Linear Method: An extension of single exponential smoothing for trend prediction. Holt-Winters No Trend Method: A variation without a trend component[@burinskiene2022forecasting]. The study focuses on short-term forecasting, excluding long-term forecasting and safety stock considerations, which are crucial in the pharmaceutical industry due to varying demand patterns[@nikolopoulos2016forecasting]. The naive method fails to capture trends or seasonality, while the moving average method struggles with sudden changes and irregular patterns[@restyana2021analysis]. Holt's linear method is limited in handling complex seasonal patterns, and the Holt-Winters method requires careful parameter tuning and historical data analysis[@koehler2001forecasting]. Single exponential smoothing is not effective for sudden changes or outliers[@inproceedings{zahra2019forecasting]. The study used MAPE, MSE, and Theil's U2 test to evaluate forecast accuracy, with Theil's U2 test comparing results to the naive forecast. A U2 value greater than 1 indicates better performance than the naive method [@burinskiene2022forecasting]. The Holt-Winters Multiplicative method was found to improve weekly forecasts and outperform other methods, especially in addressing product shortages in pharmacies.

A study conducted by Anusha et al. in the Indian pharmaceutical retail sector utilized forecasting models such as Simple Moving Average, Exponential Smoothing, and Winter’s Exponential Smoothing[@anusha2014demand]. These models were employed to predict demand for both seasonal pharmaceutical items like Okacet 10mg tablets and random pharmaceutical items like Stamlo Beta tablets. Quantitative indicators like Mean Absolute Deviation (MAD), Mean Squared Error (MSE), and Mean Absolute Percentage Error (MAPE) were used to assess forecast error. The Winter’s Exponential Smoothing (WES) model showed the least sales forecast error for the seasonal pharmaceutical item Okacet 10mg tablets, with a MAPE of 27.50, indicating high forecast accuracy. Conversely, for the non-seasonal pharmaceutical item Stamlo Beta tablets, the Six-month Moving Average model performed best with a MAPE of 45.62[@anusha2014demand]. However, the study did not account for unexplained seasonal variations in staff, shifts in the number of potential patients, physician turnover, preferences for pharmaceuticals, and the influence of marketing representatives on surrounding clinics. These factors were not factored into the analysis, potentially affecting the accuracy of demand forecasting in the pharmaceutical retail setting. These limitations suggest that the study may not have considered all relevant variables impacting demand forecasting accuracy.

Zhu et al. conducted a study comparing various forecasting methods in the pharmaceutical industry, including Moving Average (MA), Linear Regression (LR), Exponential Smoothing (ES), Exponential State Space Model (ETS), Vector Autoregressive with exogenous variables (VARX), Symbolic regression, Graph-based analysis, Artificial Neural Networks (ANN), Clustering, and Recurrent Neural Networks (RNN)[@zhu2021demand]. Additionally, machine learning algorithms such as Linear Regression (LR), Support Vector Regression (SVR), Random Forest (RF), and RNN with cross-series training were assessed.The study evaluated forecast accuracy using metrics like Normalized Mean Squared Error (NMSE), Normalized Mean Absolute Error (NMAE), and Bias. Results indicated that the cross-series training models outperformed baseline models (Moving Average, Exponential Smoothing, Linear Regression) in terms of NMSE, NMAE, and Bias. RNN models, particularly when implemented with cross-series training, showed superiority and generalizability across datasets compared to other machine learning algorithms (LR, SVR, RF), confirming the effectiveness of the proposed forecasting framework in the pharmaceutical industry.

Ribeiro et al. conducted a study aimed at detecting outliers and predicting sales for a pharmaceutical distribution company. They employed the dampened Pegels method, a time-series data mining technique, to forecast future quantities of medicines to be ordered by the company[@article{ribeiro2017improving]. The accuracy of the forecasting model was evaluated using the Symmetric Mean Absolute Percentage Error (SMAPE) metric, which indicated that the Pegels method provided more accurate sales forecasts compared to the company's previous method. Furthermore, these forecasts effectively met demand, aiding in efficient stock management and ensuring customer satisfaction while minimizing inventory costs.

The study by Bon and Ng at the University Health Centre aimed to enhance inventory demand forecasting for Panadol 650mg through the utilization of various forecasting techniques. Their research design encompassed both qualitative and quantitative data collection methods within a descriptive framework. Historical data spanning from January 2009 to August 2014 were analyzed employing methods like single moving average, exponential smoothing, and regression analysis. Forecast accuracy was assessed using metrics including RMSE, MSE, MAD, and MAPE to identify the most precise forecasting technique. The objective was to optimize inventory management by selecting the most accurate forecasting method based on these measurements[@bon2017optimization]. Results highlighted regression analysis as the superior technique among the ten methods tested, exhibiting the least forecasting error for optimizing Panadol 650mg inventory demand at the University Health Centre[@inproceedings{bon2017optimization].

The study conducted by Newberne and MIKE focused on utilizing Holt-Winters Forecasting in Healthcare, particularly in the context of healthcare management involving Pseudoephedrine prescriptions and Upper Respiratory Infection (URI) visits [@newberne2006holt]. Data from the Military Healthcare System's M2 database were analyzed using the Holt-Winters method to demonstrate its applicability in healthcare forecasting. Forecasting accuracy was assessed using the Mean Absolute Percentage Error (MAPE) metric. The study employed various forecasting methods including Naïve, Seasonal Naïve, ARIMA, and Deep Learning Model: Long Short-Term Memory (LSTM) Recurrent Neural Network. Additionally, techniques like Simple Exponential Smoothing, Holt with linear and damped trends, Holt-Winters, ES state space model, and Theta were explored to forecast drug quantities for distribution in the Brazilian Public Health System. Among these methods, the LSTM model often exhibited the best performance based on Mean Absolute Error (MAE) as a measure of forecast accuracy. However, deep learning models like LSTM may pose challenges due to their complexity and computational intensity, requiring specialized expertise for tuning and implementation in practical settings.

The study conducted by Zahra and Putra focused on using ARIMA and Single Exponential Smoothing methods to predict and calculate EOQ (Economic Order Quantity). To measure forecast accuracy, the researchers utilized RMSE, MAPE, MAD, and MSE metrics. The findings revealed that the ARIMA (1,0,0) method exhibited the smallest error values, with a MAPE of 13% and MSE of 283,715, indicating its superior accuracy compared to the Exponential Smoothing method. Based on these results, the researchers predicted a need for 222 boxes of medicine for the next period and recommended procuring 223 boxes to maintain a service level of 95%. The study underscored the significance of accurate forecasting in optimizing medicine procurement processes and ensuring adequate stock levels to efficiently meet demand. By employing MAPE and MSE as forecast accuracy metrics, the researchers could compare the performance of different forecasting methods and determine the most effective approach for predicting medicine needs in hospital settings[@zahra2019forecasting].

The study conducted by Restyana et al. focused on analyzing drug forecasting within the Jombang Regency from 2017 to 2019. They employed the Single Moving Average and Single Exponential Smoothing methods to predict drug demand across 35 public health centers. The objective was to identify the method with the lowest forecasting error by assessing metrics like Mean Absolute Deviation (MAD) and Mean Square Error (MSE). By analyzing annual primary data and testing errors, the researchers aimed to evaluate the accuracy of the forecasting methods. The findings revealed that the Single Exponential Smoothing (SES) approach exhibited superior forecast accuracy, as evidenced by lower MAD and MSE values.The study by Kim et al. used Vector Autoregressive with Exogenous Variables (VARX) models for demand forecasting of medicines. Two models were constructed to test the effect of exogenous variables: Model 1, a pure autoregressive model without exogenous variables, and Model 2, the suggested VARX model incorporating topic trends from social media analysis. The study used sales data as endogenous variables and moving averages of topic trends as exogenous variables in the VARX models.The results showed that incorporating topic trends into the VARX models enhanced the performance of demand forecasting, particularly for medicines sensitive to environmental and seasonal changes. By analyzing topic trends related to certain diseases and incorporating them into VARX models, the researchers found significant improvement in forecast accuracy.Prediction error rates were used as the forecast accuracy measure, calculated for four different medicines for each model (Model 1 and Model 2). The comparison of error rates between the two models demonstrated that adding topic trends as exogenous variables in the VARX model significantly improved the accuracy of demand forecasts, highlighting the importance of integrating social media data into forecasting processes[@kim2015demand].

Table 1 provides a summary of some studies in the literature on forecasting in pharmaceuticals products. We have identified several limitations in the existing literature on pharmaceutical product forecasting that have motivated us to pursue this research. These limitations can be summarized as follows:

1. Almost all of the studies have primarily concentrated on point forecasts at a single level.
2. A majority of the research has not explored longer time series data periods.
3. Only a few studies have integrated domain knowledge with time series data to enhance pharmaceutical forecasting accuracy.
4. Additionally, no studies have been conducted in African settings, where challenges related to forecasting and pharmaceutical availability are prevalent.

Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products

>>>>

Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes. Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed. The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid, @makridakis2020forecasting].

In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies utilizing a combination of simple and advanced methods to enhance accuracy and reliability[@rathipriya2023demand,@merkuryeva2019demand,@nikolopoulos2016forecasting,@van2021using,@khalil2014intelligent,@zhu2021demand,@anusha2014demand,@burinskiene2022forecasting,@bon2017optimization,@sousa2019statistical,@zahra2019forecasting]. Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications. In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand,@candan2014demand, and @ribeiro2017improving].

Early research in pharmaceutical forecasting primarily focused on traditional statistical methods. These methods, including ARIMA, exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting]. These models typically assess forecast accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE). For example, @newberne2006holt applied the Holt-Winters method to forecast healthcare data, focusing on prescription trends. Their findings highlighted the method's utility in short-term planning and resource management, validated through metrics like MAPE, MAD, and MSD. Similarly, @restyana2021analysis compared Single Moving Average (SMA) and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.

Recognizing the limitations of traditional models, particularly in capturing the complexities of pharmaceutical demand (e.g., seasonality, external influences), researchers began developing hybrid models. These models combine linear methods like ARIMA with nonlinear techniques such as neural networks, aiming to improve forecasting accuracy by addressing diverse data patterns. @khalil2014intelligent introduced a hybrid model that integrated ARIMA with neural networks. Their study demonstrated that this approach significantly outperformed traditional models, especially in scenarios with limited historical data. The hybrid model's effectiveness was confirmed by improvements in key performance indicators like MAPE and RMSE.

The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques. These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.@van2021using, showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy. Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE. Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting. The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories. This suggests that no single model is universally optimal, and model selection should be context-specific.

More sophisticated models, such as Neuro-Fuzzy Systems, have emerged as powerful tools for pharmaceutical forecasting. These systems combine the learning capabilities of neural networks with the reasoning abilities of fuzzy logic, offering a balanced approach that incorporates both empirical data and expert knowledge. Deep learning models, while powerful, present challenges such as the need for large datasets and substantial computational resources. Despite these challenges, models like LSTM have shown potential in capturing long-term dependencies in time series data, as demonstrated by @sousa2019statistical in their study on drug distribution in the Brazilian Public Health System. @candan2014demand employed an Adaptive Neuro-Fuzzy Inference System (ANFIS) to forecast pharmaceutical demand. Their study highlighted the system's ability to capture complex patterns in the data, resulting in highly accurate forecasts. The effectiveness of this approach was further validated through statistical tests, such as paired T-tests and mean difference analysis, which confirmed its superiority over traditional methods.

Despite the advancements in forecasting methods, several gaps remain, particularly in addressing uncertainty and incorporating domain knowledge into the modeling process. Most existing studies focus on point forecasts, which may not fully capture the range of possible outcomes in pharmaceutical demand. This limitation is critical, given the high stakes of inventory management in this industry. Moreover, there is a lack of research on the impact of integrating domain knowledge—such as insights from experienced professionals—into forecasting models. This integration could potentially improve model accuracy and reliability, especially in complex and rapidly changing environments. In summary, the literature on pharmaceutical forecasting has evolved from traditional statistical models to advanced machine learning techniques, each offering unique insights and capabilities. However, there is still room for improvement, particularly in addressing uncertainty and incorporating domain knowledge. Future research should focus on developing probabilistic forecasting models and exploring ways to integrate expert insights into the forecasting process. These advancements could significantly enhance decision-making and operational efficiency in the pharmaceutical supply chain.

# Table 1: Summary of some studies on forecasting in pharmaceutical supply chain

# How the domain knowledge were collected

To gather domain knowledge, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit. Five experts with over 10 years of experience in managing the distribution of the pharmaceuticals in EPSS agreed to participate. The researchers, along with these experts, examined the electronic bin cards of selected pharmaceuticals. A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom. The term 'bin' refers to the specific storage location for an item. Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance. In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential. It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.

We meticulously traced sales data for the past five years, triangulating graphs with bin card records to identify predictors influencing demand for each product. The expert uses their domain knowledge and experience to identify about the distribution charachterstics of the specific products and sometimes when they assume that they do not have enough history or do not remember about that specific products they call to other experts who remember specific issues and decide whetrere the specific predictos affect the distribution and if all of them agree the include it as predcitors. The data collection process took more than a month. We downloaded and analyzed over 100 bin cards per product. Dates on the bin cards were converted from the Ethiopian calendar to the Gregorian calendar for accuracy. Transactions due to storage or warehouse changes within EPSS were excluded. After cleaning the data,and discussing with the experts  we identified the following key predictors affecting distribution and sales: stock replenishment, stock outs, COVID-19, fiscal year counting, conflict, and malaria seasonality.

•	Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints. 
•	Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand. This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process. Resulted in no transactions and a decline in sales.

•	COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain. The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products. With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted. The impact of COVID-19 on sales was significant. After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs. This push was from March to August 2020. During this period, if products were not stocked out, sales continued to hubs and health facilities.

•	Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced sales. Stores closed during these periods, halting transactions. We observed increased sales before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.

•	Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters. Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises. Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas. This disruption can cause significant delays or prevent the delivery of essential medicines altogether. Internal displacements and conflicts in Ethiopia over the past five years also affected sales and distribution. We traced periods of instability and included them as predictors. The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.

•	Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified epidemic periods affecting sales: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.

The comprehensive analysis and inclusion of these predictors ensured a robust model for forecasting pharmaceutical sales and distribution, leveraging expert domain knowledge and extensive historical data analysis.
